Learning growth mechanisms of tail realistic preferential attachment models from network degree distributions

### 2025-10-03 Files/fixed_paper.tex
Fixing small issues

Amid the recent rise of interest in networks, there has been a debate on
whether most real networks are scale-free. Claiming a real network is
scale-free is equivalent to saying that the tail of the degree
distribution follows a power law, that is the fraction of nodes with
degree greater than \(k\) is proportional to \(k^{-\alpha}\), and
therefore has a regularly varying tail with tail index \(\alpha\). On
the side against the claim is \citet{Broido_2019} who compared the fits
of a power law model against that of several non-scale-free models to
nearly a thousand networks, only to find strong evidence for
scale-freeness in four percent and weak evidence in over half of the
networks, thus claiming that scale-free networks do not make up a
majority in real networks. \citet{Voitalov_2019} disagree and claim that
these networks are not nearly as rare and only appear so as a result of
an unrealistic expectation of a power law without deviations or noise.
Additional evidence of these deviations from a power law is shown by
\citet{Lee24} who demonstrate that a lot of networks are partially
scale-free, in that the body of the degree distribution is often
modelled well by a power law, while the tail is lighter than what is
implied by the body, albeit still regularly varying.

@@

The model from \citet{Barabasi99} provided the foundations for various
generalisations --- \citet{krapivsky01} considered
\(b(k) = (k+1)^\alpha\), and showed that the degree distribution is not
regularly varying (and therefore not following the power law) when
\(0<\alpha<1\), and when \(\alpha>1\) a finite number of nodes end up
with all edges after a certain point resulting in a degenerate degree
distribution. \citet{wang2022random} returns to a linear preference
function of the form \(b(k) = k+\varepsilon\) but adds the possibility
for reciprocal edges to be sent, resulting in the joint distribution of
in-degree and out-degree being multivariate regularly varying and having
the property of hidden regular variation. \citet{rudas07} followed in
the footsteps of \citet{krapivsky01}, by considering a PA tree and using
theory from continuous branching processes, deriving a limiting degree
distribution in terms of the preference function \(b(\cdot)\).
Nevertheless, research in this area tends to only focus on the
theoretical asymptotic results of network growth models with little
analysis of real networks.

@@

While \citet{Voitalov_2019} have pioneered using methods in extreme
value theory to analyse degree distributions, while their approach is
good they overlooked the inapplicability of standard tools for
continuous extremes, due to the discrete nature of degrees. A key
difficulty arises because many standard discrete distributions, such as
the Poisson, geometric, or negative binomial, do not satisfy the
conditions required to belong to a maximum domain of attraction (MDA).
\citet{shimura12} showed that a discrete distribution \(F\) must be
long-tailed,that is
\(\lim_{x\rightarrow\infty}\bar F(x+1)/\bar F(x) = 1\), in order to be
in an MDA, and introduced the notion of recoverability to the MDA, also
sometimes referred to as the discrete MDA \citep{hitz24}. Using the
theoretical guarantees of being in a discrete MDA for a discrete
distribution by \citet{shimura12}, we demonstrate how the tail of the
degree distribution is directly influenced by \(b(\cdot)\), and
subsequently propose a class of preference functions that not only imply
regularly varying degree distributions but are also tail realistic for
real networks. These analytical results enable the likelihood of the
degree distribution to be expressed in terms of the parameters of
\(b(\cdot)\), which in turn allows the underlying mechanism of the
network, assumed to grow according to GPA, to be inferred directly.

@@

\[
\bar F(k) = \lim_{t\rightarrow\infty}\frac{\sum_{x\in\Upsilon(t)}\mathbb I\left\{\text{deg}(x,\Upsilon(t)_{\downarrow x})>k\right\}}{\sum_{x\in\Upsilon(t)} 1},
\]

For the tree \(\Upsilon(t)\) defined as above, Theorem 1 from
\citet{rudas07} states that

\begin{equation}\phantomsection\label{eq-survlim}{
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]\text{d}t.
}\end{equation} for some characteristic function of the tree
\(\varphi(\cdot)\).

Using this result, and the characteristic function
\(\varphi(\Upsilon(t)) = \mathbb I\{\text{deg}(x,\Upsilon(t))>k\}\), we
can write the limiting survival as:

@@

\begin{equation}\phantomsection\label{eq-polysurv}{
\bar F(k) = \begin{cases}
\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k\le k_0,\\
\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k > k_0,
\end{cases}
}\end{equation}

with Malthusian parameter \(\lambda^*\) satisfying
\(\hat \rho(\lambda^*)=1\) where

\begin{equation}\phantomsection\label{eq-rho}{
\hat\rho(\lambda) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda + i^\alpha + \varepsilon} 
}\end{equation}

which has to be solved numerically for most parameter choices. Also,
note that \(\lambda>\beta\). See Appendix \ref{sec-proofs} for the
derivations of Equation~\ref{eq-rho}.

@@

Therefore, \begin{equation}\phantomsection\label{eq-igp-est}{
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k\le k_0,\\
\approx \left(\prod_{i=0}^{k_0}\frac{i^\alpha + \varepsilon}{\lambda^* + i^\alpha + \varepsilon}\right)\left(\frac{\beta(k-k_0)}{k_0^{\alpha}+\varepsilon+\beta} + 1\right)^{-\lambda^{*}/\beta},&k > k_0,
\end{cases}
}\end{equation} meaning that for \(k\ge k_0\) the limiting degree
distribution (for large \(k_0^\alpha\)) is approximated by the
\(\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon+\beta}{\lambda^*},k_0-1\right)\)
distribution.

@@

\[
b(k) = \begin{cases}
g(k),&k<k_0,\\
g(k_0) + \beta(k-k_0), &k\ge k_0,
\end{cases}
\] for \(\beta>0, k_0\in\mathbb N\) we have that

@@

\begin{align*}
\sum_{n=0}^\infty\frac{\Gamma(n+x)}{\Gamma(n+x+y)} &=\frac{1}{\Gamma(y)}\sum_{n=0}^\infty \text{B}(n+x,y)\\
&=\frac{1}{\Gamma(y)}\sum_{n=0}^\infty\int_0^1t^{n+x-1}(1-t)^{y-1}\text{d}t\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-1}\sum_{n=0}^\infty t^n \text{d}t\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-1}\frac{1}{1-t}\text{d}t\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-2}at\\
&=\frac{1}{\Gamma(y)}\text{B}(x,y-1)\\
&= \frac{\Gamma(x)}{(y-1)\Gamma(x+y-1)}.
\end{align*}

>>>>

Amid the recent rise of interest in networks, there has been a debate on
whether most real networks are scale-free. Claiming a real network is
scale-free is equivalent to saying that the tail of the degree
distribution follows a power law, that is the fraction of nodes with
degree greater than \(k\) is eventually proportional to \(k^{-\alpha}\),
and therefore has a regularly varying tail with tail index \(\alpha\).
On the side against the claim, that most real networks are scale-free,
is \citet{Broido_2019} who compared the fits of a power law model
against that of several non-scale-free models to nearly a thousand
networks, only to find strong evidence for scale-freeness in four
percent and weak evidence in over half of the networks, thus claiming
that scale-free networks do not make up a majority of real networks.
\citet{Voitalov_2019} disagree and claim that these networks are not
nearly as rare and only appear so as a result of an unrealistic
expectation of a power law without deviations or noise. Additional
evidence of these deviations from a power law is shown by \citet{Lee24}
who demonstrate that a lot of networks are partially scale-free, in that
the body of the degree distribution is often modelled well by a power
law, while the tail is lighter than what is implied by the body, albeit
still regularly varying.

@@

The model from \citet{Barabasi99} provided the foundations for various
generalisations --- \citet{krapivsky01} considered
\(b(k) = (k+1)^\alpha\), and showed that the tail of the degree
distribution is not regularly varying (and therefore not following the
power law) when \(0<\alpha<1\), and when \(\alpha>1\) a finite number of
nodes end up with all edges after a certain point resulting in a
degenerate degree distribution. \citet{wang2022random} return to a
linear preference function of the form \(b(k) = k+\varepsilon\) but add
the possibility for reciprocal edges to be sent, resulting in the joint
distribution of in-degree and out-degree being multivariate regularly
varying and having the property of hidden regular variation.
\citet{rudas07} followed in the footsteps of \citet{krapivsky01}, by
considering a GPA tree and using theory from continuous branching
processes, deriving a limiting degree distribution in terms of the
preference function \(b(\cdot)\). Unfortunately, research in this area
tends to only focus on the theoretical asymptotic results of network
growth models with little analysis of real networks.

@@

While \citet{Voitalov_2019} have pioneered using methods in extreme
value theory to analyse degree distributions, while their approach is
good they overlooked the inapplicability of standard tools for
continuous extremes, due to the discrete nature of degrees. A key
difficulty arises because many standard discrete distributions, such as
the Poisson, geometric, or negative binomial, do not satisfy the
conditions required to belong to a maximum domain of attraction (MDA).
\citet{shimura12} showed that a discrete distribution \(F\) must be
long-tailed, that is
\(\lim_{x\rightarrow\infty}\bar F(x+1)/\bar F(x) = 1\), in order to be
in an MDA, and introduced the notion of recoverability to the MDA, also
sometimes referred to as the discrete MDA \citep{hitz24}. Using the
theoretical guarantees of being in a discrete MDA for a discrete
distribution by \citet{shimura12}, we demonstrate how the tail of the
degree distribution is directly influenced by \(b(\cdot)\), and
subsequently propose a class of preference functions that not only imply
regularly varying degree distributions but are also tail realistic for
real networks. These analytical results enable the likelihood of the
degree distribution to be expressed in terms of the parameters of
\(b(\cdot)\), which in turn allows the underlying mechanism of the
network, assumed to grow according to GPA, to be inferred directly.

@@

\[
\bar F(k) = \lim_{t\rightarrow\infty}\frac{\sum_{x\in\Upsilon(t)}\mathbb I\left\{\text{deg}(x,\Upsilon(t)_{\downarrow x})>k\right\}}{\sum_{x\in\Upsilon(t)} 1},
\] for the tree \(\Upsilon(t)\) defined as above, Theorem 1 from
\citet{rudas07} states that

\begin{equation}\phantomsection\label{eq-survlim}{
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]\text{d}t.
}\end{equation}

Using this result, and the characteristic function
\(\varphi(\Upsilon(t)_{\downarrow x}) = \mathbb I\{\text{deg}(x,\Upsilon(t)_{\downarrow x})>k\}\),
we can write the limiting survival as:

@@

\begin{equation}\phantomsection\label{eq-polysurv}{
\bar F(k) = \begin{cases}
\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k\le k_0,\\
\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k > k_0,
\end{cases}
}\end{equation} with Malthusian parameter \(\lambda^*\) satisfying
\(\hat \rho(\lambda^*)=1\) where

\begin{equation}\phantomsection\label{eq-rho}{
\hat\rho(\lambda) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda + i^\alpha + \varepsilon}, 
}\end{equation} which has to be solved numerically for most parameter
choices. Also, note that \(\lambda>\beta\). See Appendix
\ref{sec-proofs} for the derivations of Equation~\ref{eq-rho}.

@@

Therefore, \begin{equation}\phantomsection\label{eq-igp-est}{
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k\le k_0,\\
\approx \left(\prod_{i=0}^{k_0}\frac{i^\alpha + \varepsilon}{\lambda^* + i^\alpha + \varepsilon}\right)\left(\frac{\beta(k-k_0)}{k_0^{\alpha}+\varepsilon+\beta} + 1\right)^{-\lambda^{*}/\beta},&k > k_0,
\end{cases}
}\end{equation} meaning that for \(k> k_0\) the limiting degree
distribution (for large \(k_0^\alpha\)) is approximated by the
\(\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon+\beta}{\lambda^*},k_0\right)\)
distribution.

@@

\[
b(k) = \begin{cases}
g(k),&k\le k_0,\\
g(k_0) + \beta(k-k_0), &k> k_0,
\end{cases}
\] for \(\beta>0, k_0\in\mathbb N\) we have that

@@

\begin{align*}
\sum_{n=0}^\infty\frac{\Gamma(n+x)}{\Gamma(n+x+y)} &=\frac{1}{\Gamma(y)}\sum_{n=0}^\infty \text{B}(n+x,y)\\
&=\frac{1}{\Gamma(y)}\sum_{n=0}^\infty\int_0^1t^{n+x-1}(1-t)^{y-1}\text{d}t\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-1}\sum_{n=0}^\infty t^n \text{d}t\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-1}\frac{1}{1-t}\text{d}t\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-2}\text{d}t\\
&=\frac{1}{\Gamma(y)}\text{B}(x,y-1)\\
&= \frac{\Gamma(x)}{(y-1)\Gamma(x+y-1)}.
\end{align*}

### 2025-10-03 Files/paper.qmd
Updating pdf
Fixing small issues

---
title: "Learning growth mechanisms of tail realistic preferential attachment models from network degree distributions"
crossref:
  fig-title: '**Fig.**'        
format:
  nature-pdf:
    fontsize: 10pt
    indent: true
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-basic
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    number-sections: true
author:
  - name: Thomas William Boughen
    orcid: 0009-0005-1979-2479
    email: t.w.boughen1@newcastle.ac.uk
    corresponding: true
    affiliations:
      - name: Newcastle University
        id: 1
        department: School of Mathematics, Statistics and Physics
  - name: Clement Lee
    affiliations:
      - ref: 1
  - name: Vianey Palacios Ramírez
    affiliations:
      - ref: 1
abstract: |
  Identifying the generating mechanism of a network is challenging as, more often than not, only snapshots are available, but not the full evolution. One candidate for the generating mechanism is preferential attachment which, in its simplest form, results in a degree distribution that follows the power law. Consequently, the growth of real-life networks that display such power-law behaviour is commonly modelled by preferential attachment. The ubiquity of the power law has been challenged by the presence of alternatives with comparable performance, as well as the recent findings that the tail of the degree distribution is often lighter than implied by the body, whilst still being regularly varying. In this paper, we propose a preferential attachment model with a flexible preference function. Using methods for discrete extremes, we characterise the tail behaviour of the limiting degree distribution directly by the form of the preference function. Directly relating the tail index to the model parameters enables them to be inferred when fitting to degree distributions alone, which is supported by simulation studies. Results of applications to real data are promising and comparable to alternatives.
keywords: Extremes · Discrete extremes · Power law · Preferential attachment
msc: 
 - 05C82
 - 60G70
bibliography: refs.bib  
keep-tex: true
---

@@

Amid the recent rise of interest in networks, there has been a debate on whether most real networks are scale-free. Claiming a real network is scale-free is equivalent to saying that the tail of the degree distribution follows a power law, that is the fraction of nodes with degree greater than $k$ is proportional to $k^{-\alpha}$, and therefore has a regularly varying tail with tail index $\alpha$. On the side against the claim is @Broido_2019 who compared the fits of a power law model against that of several non-scale-free models to nearly a thousand networks, only to find strong evidence for scale-freeness in four percent and weak evidence in over half of the networks, thus claiming that scale-free networks do not make up a majority in real networks. @Voitalov_2019 disagree and claim that these networks are not nearly as rare and only appear so as a result of an unrealistic expectation of a power law without deviations or noise. Additional evidence of these deviations from a power law is shown by @Lee24 who demonstrate that a lot of networks are partially scale-free, in that the body of the degree distribution is often modelled well by a power law, while the tail is lighter than what is implied by the body, albeit still regularly varying.

@@

The model from @Barabasi99 provided the foundations for various generalisations — @krapivsky01 considered $b(k) = (k+1)^\alpha$, and showed that the degree distribution is not regularly varying (and therefore not following the power law) when $0<\alpha<1$, and when $\alpha>1$ a finite number of nodes end up with all edges after a certain point resulting in a degenerate degree distribution. @wang2022random returns to a linear preference function of the form $b(k) = k+\varepsilon$ but adds the possibility for reciprocal edges to be sent, resulting in the joint distribution of in-degree and out-degree being multivariate regularly varying and having the property of hidden regular variation. @rudas07 followed in the footsteps of @krapivsky01, by considering a PA tree and using theory from continuous branching processes, deriving a limiting degree distribution in terms of the preference function $b(\cdot)$. Nevertheless, research in this area tends to only focus on the theoretical asymptotic results of network growth models with little analysis of real networks.

@@

While @Voitalov_2019 have pioneered using methods in extreme value theory to analyse degree distributions, while their approach is good they overlooked the inapplicability of standard tools for continuous extremes, due to the discrete nature of degrees. A key difficulty arises because many standard discrete distributions, such as the Poisson, geometric, or negative binomial, do not satisfy the conditions required to belong to a maximum domain of attraction (MDA). @shimura12 showed that a discrete distribution $F$ must be long-tailed,that is $\lim_{x\rightarrow\infty}\bar F(x+1)/\bar F(x) = 1$, in order to be in an MDA, and introduced the notion of recoverability to the MDA, also sometimes referred to as the discrete MDA [@hitz24]. Using the theoretical guarantees of being in a discrete MDA for a discrete distribution by @shimura12, we demonstrate how the tail of the degree distribution is directly influenced by $b(\cdot)$, and subsequently propose a class of preference functions that not only imply regularly varying degree distributions but are also tail realistic for real networks. These analytical results enable the likelihood of the degree distribution to be expressed in terms of the parameters of $b(\cdot)$, which in turn allows the underlying mechanism of the network, assumed to grow according to GPA, to be inferred directly.

@@

$$
\bar F(k) = \lim_{t\rightarrow\infty}\frac{\sum_{x\in\Upsilon(t)}\mathbb I\left\{\text{deg}(x,\Upsilon(t)_{\downarrow x})>k\right\}}{\sum_{x\in\Upsilon(t)} 1},
$$

For the tree $\Upsilon(t)$ defined as above, Theorem 1 from @rudas07 states that

$$
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]\text{d}t.
$$ {#eq-survlim}
for some characteristic function of the tree $\varphi(\cdot)$.

Using this result, and the characteristic function $\varphi(\Upsilon(t)) = \mathbb I\{\text{deg}(x,\Upsilon(t))>k\}$, we can write the limiting survival as:

@@

$$
\bar F(k) = \begin{cases}
\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k\le k_0,\\
\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k > k_0,
\end{cases}
$$ {#eq-polysurv}

with Malthusian parameter $\lambda^*$ satisfying $\hat \rho(\lambda^*)=1$ where

$$
\hat\rho(\lambda) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda + i^\alpha + \varepsilon} 
$$ {#eq-rho}

which has to be solved numerically for most parameter choices. Also, note that $\lambda>\beta$. See Appendix [-@sec-proofs] for the derivations of @eq-rho.

@@

Therefore, $$
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k\le k_0,\\
\approx \left(\prod_{i=0}^{k_0}\frac{i^\alpha + \varepsilon}{\lambda^* + i^\alpha + \varepsilon}\right)\left(\frac{\beta(k-k_0)}{k_0^{\alpha}+\varepsilon+\beta} + 1\right)^{-\lambda^{*}/\beta},&k > k_0,
\end{cases}
$$ {#eq-igp-est} meaning that for $k\ge k_0$ the limiting degree distribution (for large $k_0^\alpha$) is approximated by the $\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon+\beta}{\lambda^*},k_0-1\right)$ distribution.

@@

$$
b(k) = \begin{cases}
g(k),&k<k_0,\\
g(k_0) + \beta(k-k_0), &k\ge k_0,
\end{cases}
$$ for $\beta>0, k_0\in\mathbb N$ we have that

>>>>

---
title: "Learning growth mechanisms of tail realistic preferential attachment models from network degree distributions"
crossref:
  fig-title: '**Fig.**'        
format:
  nature-pdf:
    fontsize: 10pt
    indent: true
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-basic
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    number-sections: true
author:
  - name: Thomas William Boughen
    orcid: 0009-0005-1979-2479
    email: t.w.boughen1@newcastle.ac.uk
    corresponding: true
    affiliations:
      - name: Newcastle University
        id: 1
        department: School of Mathematics, Statistics and Physics
  - name: Clement Lee
    affiliations:
      - ref: 1
  - name: Vianey Palacios Ramírez
    affiliations:
      - ref: 1
abstract: |
  Identifying the generating mechanism of a network is challenging as, more often than not, only snapshots are available, but not the full evolution. One candidate for the generating mechanism is preferential attachment which, in its simplest form, results in a degree distribution that follows the power law. Consequently, the growth of real-life networks that display such power-law behaviour is commonly modelled by preferential attachment. The ubiquity of the power law has been challenged by the presence of alternatives with comparable performance, as well as the recent findings that the tail of the degree distribution is often lighter than implied by the body, whilst still being regularly varying. In this paper, we propose a preferential attachment model with a flexible preference function. Using methods for discrete extremes, we characterise the tail behaviour of the limiting degree distribution directly by the form of the preference function. Directly relating the tail index to the model parameters enables them to be inferred when fitting to degree distributions alone, which is supported by simulation studies. Results of applications to real data are promising and comparable to alternatives.
keywords: Extremes · Discrete extremes · Power law · Preferential attachment
msc: 
 - 05C82
 - 60G70
bibliography: refs.bib  
keep-tex: true
pdf-engine: xelatex
---

@@

Amid the recent rise of interest in networks, there has been a debate on whether most real networks are scale-free. Claiming a real network is scale-free is equivalent to saying that the tail of the degree distribution follows a power law, that is the fraction of nodes with degree greater than $k$ is eventually proportional to $k^{-\alpha}$, and therefore has a regularly varying tail with tail index $\alpha$. On the side against the claim, that most real networks are scale-free, is @Broido_2019 who compared the fits of a power law model against that of several non-scale-free models to nearly a thousand networks, only to find strong evidence for scale-freeness in four percent and weak evidence in over half of the networks, thus claiming that scale-free networks do not make up a majority of real networks. @Voitalov_2019 disagree and claim that these networks are not nearly as rare and only appear so as a result of an unrealistic expectation of a power law without deviations or noise. Additional evidence of these deviations from a power law is shown by @Lee24 who demonstrate that a lot of networks are partially scale-free, in that the body of the degree distribution is often modelled well by a power law, while the tail is lighter than what is implied by the body, albeit still regularly varying.

@@

The model from @Barabasi99 provided the foundations for various generalisations — @krapivsky01 considered $b(k) = (k+1)^\alpha$, and showed that the tail of the degree distribution is not regularly varying (and therefore not following the power law) when $0<\alpha<1$, and when $\alpha>1$ a finite number of nodes end up with all edges after a certain point resulting in a degenerate degree distribution. @wang2022random return to a linear preference function of the form $b(k) = k+\varepsilon$ but add the possibility for reciprocal edges to be sent, resulting in the joint distribution of in-degree and out-degree being multivariate regularly varying and having the property of hidden regular variation. @rudas07 followed in the footsteps of @krapivsky01, by considering a GPA tree and using theory from continuous branching processes, deriving a limiting degree distribution in terms of the preference function $b(\cdot)$. Unfortunately, research in this area tends to only focus on the theoretical asymptotic results of network growth models with little analysis of real networks.

@@

While @Voitalov_2019 have pioneered using methods in extreme value theory to analyse degree distributions, while their approach is good they overlooked the inapplicability of standard tools for continuous extremes, due to the discrete nature of degrees. A key difficulty arises because many standard discrete distributions, such as the Poisson, geometric, or negative binomial, do not satisfy the conditions required to belong to a maximum domain of attraction (MDA). @shimura12 showed that a discrete distribution $F$ must be long-tailed, that is $\lim_{x\rightarrow\infty}\bar F(x+1)/\bar F(x) = 1$, in order to be in an MDA, and introduced the notion of recoverability to the MDA, also sometimes referred to as the discrete MDA [@hitz24]. Using the theoretical guarantees of being in a discrete MDA for a discrete distribution by @shimura12, we demonstrate how the tail of the degree distribution is directly influenced by $b(\cdot)$, and subsequently propose a class of preference functions that not only imply regularly varying degree distributions but are also tail realistic for real networks. These analytical results enable the likelihood of the degree distribution to be expressed in terms of the parameters of $b(\cdot)$, which in turn allows the underlying mechanism of the network, assumed to grow according to GPA, to be inferred directly.

@@

$$
\bar F(k) = \lim_{t\rightarrow\infty}\frac{\sum_{x\in\Upsilon(t)}\mathbb I\left\{\text{deg}(x,\Upsilon(t)_{\downarrow x})>k\right\}}{\sum_{x\in\Upsilon(t)} 1},
$$
for the tree $\Upsilon(t)$ defined as above, Theorem 1 from @rudas07 states that

$$
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]\text{d}t.
$$ {#eq-survlim}

Using this result, and the characteristic function $\varphi(\Upsilon(t)_{\downarrow x}) = \mathbb I\{\text{deg}(x,\Upsilon(t)_{\downarrow x})>k\}$, we can write the limiting survival as:

@@

$$
\bar F(k) = \begin{cases}
\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k\le k_0,\\
\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k > k_0,
\end{cases}
$$ {#eq-polysurv}
with Malthusian parameter $\lambda^*$ satisfying $\hat \rho(\lambda^*)=1$ where

$$
\hat\rho(\lambda) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda + i^\alpha + \varepsilon}, 
$$ {#eq-rho}
which has to be solved numerically for most parameter choices. Also, note that $\lambda>\beta$. See Appendix [-@sec-proofs] for the derivations of @eq-rho.

@@

Therefore, $$
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k\le k_0,\\
\approx \left(\prod_{i=0}^{k_0}\frac{i^\alpha + \varepsilon}{\lambda^* + i^\alpha + \varepsilon}\right)\left(\frac{\beta(k-k_0)}{k_0^{\alpha}+\varepsilon+\beta} + 1\right)^{-\lambda^{*}/\beta},&k > k_0,
\end{cases}
$$ {#eq-igp-est} meaning that for $k> k_0$ the limiting degree distribution (for large $k_0^\alpha$) is approximated by the $\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon+\beta}{\lambda^*},k_0\right)$ distribution.

@@

$$
b(k) = \begin{cases}
g(k),&k\le k_0,\\
g(k_0) + \beta(k-k_0), &k> k_0,
\end{cases}
$$ for $\beta>0, k_0\in\mathbb N$ we have that
